---
title: "Introduction to ggplot2"
author: "Jean-Baptiste Lecomte"
output: slidy_presentation
---

```{r  setup, include=FALSE}
library(knitr)
opts_chunk$set(fig.path='figure/', fig.align='center', fig.show='hold',
               size='footnotesize', out.width='.9\\linewidth')
opts_template$set(
  fig.fullpage = list(fig.width = 8, fig.height = 12),
  fig.large_square = list(fig.width = 8, fig.height = 8),
  fig.large = list(fig.width = 8, fig.height = 5),
  fig.medium = list(fig.width = 8, fig.height = 4.5),
  fig.small = list(fig.width = 8, fig.height = 3)
)
```


## Introduction

- developed by Hadley Wickham (Rice University, Houston, USA)
- highly recommended R packages to work with ggplot2: reshape2 and plyr (also developed by H. Wickham)
- first version called in 2007



<!-- ## Useful books} -->
<!-- \begin{columns} -->
<!-- \begin{column}{0.5\textwidth} -->
<!-- \includegraphics[scale=0.45]{r graphicscookbook} -->
<!-- \end{column} -->

<!-- \begin{column}{0.5\textwidth} -->
<!-- \includegraphics[scale=0.45]{ggplot2} -->
<!-- \end{column} -->
<!-- \end{columns} -->


## Online resources

- ggplot2 official documentation:\\ <http://docs.ggplot2.org/current/>
- R code related to ggplot2 cookbook:\\<http://www.cookbook-r.com/Graphs/>
- R code related to useR! ggplot2 book:\\<http://ggplot2.org/book/>
- Google groups to ask questions:\\<ggplot2```googlegroups.com>
- Statistical tools for high-throughput data analysis:\\<http://www.sthda.com/english/wiki/ggplot2-essentials>
- Github repository:\\<https://github.com/yhat/ggplot/>
- RStudio Cheat Sheet :\\<http://www.rstudio.com/wp-content/uploads/2015/12/ggplot2-cheatsheet-2.0.pdf>




## Introduction

- based on *new* aesthetic principles
- based on *The grammar of graphics* developed by Wilkinson in 2005
- efficient way to produce simple graphics with a length reduction of R code


Forget about R base graphics:
*\texttt{plot(), hist(), par(), layout(), points(), lines(),legend()*



## Principle
ggplot2 is based on a **layer** system which can be used as objects.\\
Main layers

- data $\rightarrow$ raw data
- mapping $\rightarrow$ graphic projection
- geom $\rightarrow$ geometric objects (points, lines, polygons, ...)
- stat $\rightarrow$ statistics transformation (histogram, model)
- scale $\rightarrow$ aesthetics customization (colour, shape, size, axes, legend)
- coord $\rightarrow$ coordinate system (axes, grid)
- facet $\rightarrow$ subdivision (lattice, trellis)



## Base functions
ggplot2 is based on two functions:


-  *qplot()* for **q**uick **plot**

- easy and fast, but too simple in most cases
 *qplot(x, y, data=data)*


- *ggplot()*

- more complex but more powerful and flexible by adding *layers
- *ggplot(data=data, aes(x, y)) + layers*

## Getting Started

Always work with a data.frame

Our data frame is based on the Hecate Strait Synoptic Trawl Survey and simulated data.

Download all the codes from a Github repository: 
<https://github.com/JBLecomte/ggplot2-Introduction.git>


## Getting Started
```{r str data, size = "scriptsize"}
str(df_data)
```

## A quick overview of the ggplot2 types

- Points, as for a scatterplot: *geom_point()*
- Lines: *geom_line()*
- Histogram: *geom_freqpoly(geom_histogram(), stat_bin())*
- Boxplot: *geom_boxplot()*
- Polygon: *geom_polygon()*
- Draw rectangles: *geom_raster()*
- Smooth density estimate: *geom_density()*
- Ribbons and area plots: *geom_ribbon(geom_area())*
- Map: *ggmap()*


## Scatter plot: Depth and Biomass
```{r sp_base, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp <- ggplot(data=df_data, aes(x=Avg_net_depth, y=Biomass)) +
  geom_point(size=1)
print(sp)
```


## Scatter plot with color: Depth and Biomass
```{r sp_color, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_color <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                                color=AREA)) +
  geom_point(size=1)
print(sp_color)
```



## Scatter plot with shape: Depth and Biomass
```{r sp_shape, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_shape <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                                color=AREA, shape=Year_fac)) +
  geom_point(size=1)
print(sp_shape)
```


## Scatter plot with continuous color: Depth and Biomass
```{r sp_color_cont, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_color_cont <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                                     color=Avg_net_temp)) +
  geom_point(size=1)
print(sp_color_cont)
```



## Scatter plot with size: Depth and Biomass
```{r sp_color_size, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_area <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                               size=nFish)) +
  geom_point(shape=17)
print(sp_area)
```


## Improvement of a plot
```{r sp_shape2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_shape <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                                color=AREA, shape=Year_fac)) +
  geom_point(size=1)
print(sp_shape)
```


## Improvement of a plot: axes names
```{r sp_shape_imp1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_shape_imp1 <- sp_shape + 
  xlab('Average net depth (m)') +
  ylab('Biomass (kg)')
print(sp_shape_imp1)
```


## Improvement of a plot: axes options
```{r scale, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
scale_x_continuous(name, breaks, labels, limits, trans)
scale_y_continuous(name, breaks, labels, limits, trans)
```


-    name : x or y axis labels
-    breaks : to control the breaks in the guide (axis ticks, grid lines,...).
Among the possible values, there are :
NULL : hide all breaks
waiver() : the default break computation a character or numeric vector specifying the breaks to display
- labels : labels of axis tick marks. Allowed values are :
NULL for no labels
waiver() for the default labels
character vector to be used for break labels
- limits : a numeric vector specifying x or y axis limits (min, max)
- trans for axis transformations. Possible values are "log2", "log10"



## Improvement of a plot: axes options
```{r sp_shape_imp1x, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_shape_imp1 <- sp_shape_imp1 + 
  scale_x_continuous(name='Average net depth (m)',
                     breaks=seq(-4,7,2), limits=c(-4,7))
print(sp_shape_imp1)
```


## Improvement of a plot: legend titles
```{r sp_shape_imp2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_shape_imp2 <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                                     color=AREA, shape=Year_fac)) +
  geom_point(size=1) + xlab('Average net depth (m)') + 
  scale_shape_discrete(name="Years") + scale_color_discrete(name="Area")
print(sp_shape_imp2)
```



## Improvement of a plot: plot title
```{r sp_shape_imp3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_shape_imp3 <- sp_shape_imp2 + 
  ggtitle("Biomass of species X and average net depth")
print(sp_shape_imp3)
```



## Discrete color scale
```{r sp_discrete_color1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c <- ggplot(df_data, aes(x=Avg_net_depth, y=Biomass,
                            color=Year_fac)) +
  geom_point(alpha=0.8, size=1) + xlab('Average net depth (m)')
print(sp_c)
```


## Discrete color scale: *scale_colour_hue*
```{r sp_discrete_color_hue1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Adjust luminosity and chroma
sp_c_hue <- sp_c +
  scale_colour_hue(name='Year', l=70, c=150)
print(sp_c_hue)
```


## Discrete color scale: *scale_colour_hue*
```{r sp_discrete_color_hue2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Adjust luminosity and chroma
sp_c_hue <- sp_c +
  scale_colour_hue(name='Year', l=10, c=150)
print(sp_c_hue)
```


## Discrete color scale: *scale_colour_hue*
```{r sp_discrete_color_hue2b, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Adjust luminosity and chroma
sp_c_hue <- sp_c +
  scale_colour_hue(name='Year', l=70, c=50)
print(sp_c_hue)
```


## Discrete color scale: *scale_colour_hue*
```{r sp_discrete_color_hue3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Change range of hues used
sp_c_hue <- sp_c +
  scale_colour_hue(name='Year', h=c(0, 90))
print(sp_c_hue)
```


## Discrete color scale: *scale_colour_hue*
```{r sp_discrete_color_hue3b, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Change range of hues used
sp_c_hue <- sp_c +
  scale_colour_hue(name='Year', h=c(100, 300))
print(sp_c_hue)
```


## Discrete color scale: brewer palettes


<!-- \includegraphics[scale=0.35]{colorbrewer_names.png} -->



## Discrete color scale: brewer palette
```{r sp_discrete_color4, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c_brewer <- sp_c +
  scale_color_brewer(name='Year', palette="Dark2")
print(sp_c_brewer)
```



## Discrete color scale: brewer palette
```{r sp_discrete_color4b, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c_brewer <- sp_c +
  scale_color_brewer(name='Year', palette="Set1")
print(sp_c_brewer)
```


## Discrete color scale: wesanderson palettes
```{r wesanderson, eval=FALSE, echo=TRUE}
library(wesanderson)
```

<!-- \includegraphics[scale=0.5]{wesanderson-color.png} -->


Other resources: 

<http://www.stat.columbia.edu/~tzheng/files/Rcolor.pdf>
<http://research.stowers-institute.org/efg/R/Color/Chart/ColorChart.pdf>




## Discrete color scale: wesanderson palettes
```{r sp_discrete_color_wa2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c_wanderson <- sp_c +
  scale_color_manual(name='Year', values=wes_palette(name="Darjeeling1"))
print(sp_c_wanderson)
```

## Discrete color scale: *scale_color_grey*
```{r sp_discrete_color5, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c_grey <- sp_c +
  scale_color_grey(name='Year')
print(sp_c_grey)
```



## Discrete color scale: *scale_color_grey*
```{r sp_discrete_color6, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c_grey <- sp_c +
  scale_color_grey(name='Year', start=0.8, end=0.2)
print(sp_c_grey)
```


## Discrete color scale: manual
```{r sp_discrete_color2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_c_manual <- sp_color +
  scale_color_manual(name='Area', values=c("#E69F00", "#56B4E9"))
print(sp_c_manual)
```





## Color scale: make your own
A color-blind palette:
```{r cb_palette, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
cb_palette_black <- c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                      "#0072B2", "#D55E00", "#CC79A7", "#000000")
```
Or create your own gradient palette function with 2 or more colors:
```{r Palette_grb, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
## Palette from white-grey to red to black
Palette_grb <- colorRampPalette(c(colors()[139], # green
                                  colors()[554], # red
                                  colors()[153]) # black
                )
```


## Discrete color scale: manual
```{r sp_discrete_colorcb, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_color_cb <- sp_c +
  scale_color_manual(name='Year', values = cb_palette_black)
print(sp_color_cb)
```


## Discrete color scale: manual
```{r sp_discrete_colorown, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_color_own <- sp_c +
  scale_color_manual(name='Year', values = Palette_grb(nYear))
print(sp_color_own)
```


## Continuous color scale
```{r sp_continuous_color1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc <- ggplot(df_data, aes(x=Avg_net_temp, y=Avg_net_depth)) +
  geom_point(alpha=0.8, size=1, aes(colour=Biomass)) + 
  xlab('Average net temperature (C)') + ylab('Average net depth (m)')
print(sp_cc)
```


## Continuous color scale: choose color
```{r sp_continuous_color2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc2 <- sp_cc +
  scale_colour_gradient(low = 'blue', high = 'red')
print(sp_cc2)
```


## Continuous color scale: divergent scale
```{r sp_continuous_color3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc_div <- sp_cc +
  scale_colour_gradient2(midpoint = 100,
                         low = 'blue', high = 'red', mid = muted('green'))
print(sp_cc_div)
```


## Continuous color scale: rainbow scale
```{r sp_continuous_color4, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc_own <- sp_cc +
  scale_color_gradientn(colors = rainbow(5))
print(sp_cc_own)
```


## Continuous color scale: wesanderson scale
```{r sp_continuous_color5, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc_own <- sp_cc +
  scale_color_gradientn(colors = wes_palette(n=5,
                                             name="Rushmore",
                                             type = 'continuous'))
print(sp_cc_own)
```


## Continuous color scale: own scale
```{r sp_continuous_color6, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc_own <- sp_cc +
  scale_color_gradientn(colors = Palette_grb(10))
print(sp_cc_own)
```


## Continuous color scale: own scale
```{r sp_continuous_color7, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_cc_own <- sp_cc +
  scale_color_gradientn(colors = Palette_grb(3))
print(sp_cc_own)
```


## Time series
```{r sp_TSplot, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot <- ggplot(data=df_data, aes(x=Date, y=Biomass)) + 
  geom_point() + ylim(0,max(df_data$Biomass))
print(TSplot)
```


## Time series
```{r sp_TSplot_year, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_year <- ggplot(data=df_data, aes(x=Year, y=Biomass)) + 
  geom_point()
print(TSplot_year)
```



## Time series: summarize
Summarize the data with the mean, median and quantiles for easy plotting.
The ddply function from the plyr R package provides an easy way to summarize the data.
```{r summary_data_year, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
df_summary <- ddply(df_data, c('Year'), summarise,
                    B_mean=mean(Biomass),
                    B_median=median(Biomass),
                    B_sd=sd(Biomass),
                    B_q025=quantile(Biomass, probs = 0.025),
                    B_q05=quantile(Biomass, probs = 0.05),
                    B_q10=quantile(Biomass, probs = 0.10),
                    B_q50=quantile(Biomass, probs = 0.50),
                    B_q90=quantile(Biomass, probs = 0.90),
                    B_q95=quantile(Biomass, probs = 0.95),
                    B_q975=quantile(Biomass, probs = 0.975))
```


## Time series: summarize
Summarize the data with the mean, median and quantiles for easy plotting.
The ddply function from the plyr R package provides an easy way to summarize the data.
```{r summary_data_year2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
str(df_summary)
```

## Time series with error bars
```{r sp_TSplot_i95a, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_i95 <- ggplot(data=df_summary, aes(x=Year, y=B_mean)) + 
  geom_point()
print(TSplot_i95)
```


## Time series with error bars
```{r sp_TSplot_i95b, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_i95 <- TSplot_i95 +
  geom_pointrange(aes(ymin = B_q025, ymax = B_q975))
print(TSplot_i95)
```



## Time series with error bars
```{r sp_TSplot_errori95, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_errori95 <- ggplot(data=df_summary, aes(x=Year, y=B_mean)) + 
  geom_line() + 
  geom_errorbar(aes(ymin = B_q025, ymax = B_q975), width = 0.2)
print(TSplot_errori95)
```




## Time series with error bars: Improvement
```{r sp_TSplot_i95_bis, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_i95 <- ggplot(data=df_summary, aes(x=Year, y=B_mean)) + 
  geom_point() + 
  geom_pointrange(aes(ymin = B_q025, ymax = B_q975))
print(TSplot_i95)
```


## Time series with error bars: Improvement
```{r sp_TSplot_i95_imp, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_i95 <- ggplot(data=df_summary, aes(x=Year, y=B_mean)) + 
  geom_point() + geom_pointrange(aes(ymin = B_q025, ymax = B_q975)) +
  ylab('Biomass') +
  scale_x_continuous(name = 'Year', breaks = seq(2005, 2013, by = 1))
print(TSplot_i95)
```


## Time series with *geom_ribbon*
```{r sp_TSplot_ri95, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot_ri95 <- ggplot(data=df_summary, aes(x=Year, y=B_mean)) + 
  geom_ribbon(aes(ymin = B_q025, ymax = B_q975), fill = "blue", alpha=0.2) +
  geom_point() + geom_line() + ylab('Biomass') +
  scale_x_continuous(name = 'Year', breaks = seq(2005, 2013, by = 1))
print(TSplot_ri95)
```

## Time series: x-axis as date
```{r TSplot_i95d1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
TSplot <- ggplot(data=df_data, aes(x=Date, y=Biomass)) + 
  geom_point() + 
  ylab('Biomass (kg)')
print(TSplot)
```


## Time series: x-axis as date
```{r TSplot_i95d4, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Months only
TSplotm <- TSplot +
  scale_x_date(labels = date_format("%y/%b"), date_breaks="1 year",
               date_minor_breaks = "1 month") +
  theme(axis.text.x = element_text(angle=45))
print(TSplotm)
```


## Time series: x-axis as date
```{r TSplot_i95d3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
# Format : Year/month
TSplotym <- TSplot +
  scale_x_date(labels = date_format("%Y/%m"))
print(TSplotym)
```


## Time series: x-axis as date
```{r TSplot_i95d2, eval=TRUE,  tidy=FALSE,warning=FALSE, opts.label='fig.medium'}
# Format : Year/month/day
lmin <- as.Date("2011-1-1"); lmax <- max(df_data$Date)
TSplotymdl <- TSplot +
  scale_x_date(labels = date_format("%y/%m/%d"), limits = c(lmin, lmax))
print(TSplotymdl)
```


## Boxplot
```{r boxplot_TS, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
boxplot_TS <- ggplot(data=df_data, aes(x=Year, y=Biomass)) + 
  geom_boxplot()
print(boxplot_TS)
```



## Boxplot with Year as a factor
```{r boxplot_TSf, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
boxplot_TSf <- ggplot(data=df_data, aes(x=factor(Year), y=Biomass)) + 
  geom_boxplot() 
print(boxplot_TSf)
```


## Boxplot with Year and Area as a factor
```{r boxplot_TSf_AREA, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
boxplot_TS_AREA <- ggplot(data=df_data, aes(x=factor(Year), y=Biomass, 
                                            colour=AREA)) +
  geom_boxplot()
print(boxplot_TS_AREA)
```



## Boxplot: x-axis changing label
```{r boxplot_TSf_AREA_xaxis, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
boxplot_TS_AREA <- boxplot_TS_AREA +
  scale_x_discrete(name=NULL,
                   breaks=c("2005", "2009", "2013"),
                   labels=c("Year-2005", "Year-2009", "Year-2013"))
print(boxplot_TS_AREA)
```




## Boxplot (and all other plots): changing label in the legend
The easiest way to change the label in the legend (i.e. 5AB and 5CD) rename the factor levels before the plot:

```{r renaming_factor, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
df_data$AREA <- factor(df_data$AREA, levels=c("5AB", "5CD"),
                       labels=c("Northern area", "Southern area"))
```

This code can also be useful for reordering the factor levels before the plot:
```{r reordering_factor, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
df_data$AREA <- factor(df_data$AREA, levels=c("5CD", "5AB"),
                       labels=c("Southern area", "Northern area"))
```


## Boxplot with changed factor label
```{r boxplot_TSf_AREA_final, eval=TRUE,  tidy=FALSE, opts.label='fig.small'}
bplot_TS <- ggplot(df_data, aes(x=factor(Year), y=Biomass, colour=AREA)) +
  geom_boxplot() + 
  scale_x_discrete(name=NULL,
                   breaks=c("2005", "2009", "2013"),
                   labels=c("Year-2005", "Year-2009", "Year-2013")) +
  scale_color_manual(name='Area', values = cb_palette_black) +
  theme(legend.position=c(0.2,0.9))
print(bplot_TS)
```
```{r renaming_factor_back, eval=TRUE, echo=FALSE, tidy=FALSE}
df_data$AREA <- factor(df_data$AREA, labels=c("5AB", "5CD"),
                       levels=c("Northern area", "Southern area"))
```


## Faceting: *facet_wrap*
```{r facet_wrap_0, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
hist <- ggplot(data=df_data, aes(x=DURATION_MINUTES, fill=AREA)) +
  geom_histogram()
print(hist)
```


## Faceting: *facet_wrap*
```{r facet_wrap_1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fw1 <- ggplot(data=df_data, aes(x=DURATION_MINUTES)) +
  geom_histogram(binwidth=1) + facet_wrap(~ AREA)
print(fw1)
```


## *facet_wrap*: free scales
```{r facet_wrap_2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fw1_free <- ggplot(data=df_data, aes(x=DURATION_MINUTES)) +
  geom_histogram(binwidth=1) + facet_wrap(~ AREA , scales = 'free')
print(fw1_free)
```


## *facet_wrap*: free y scale
```{r facet_wrap_3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fw1_free_y <- ggplot(data=df_data, aes(x=DURATION_MINUTES)) +
  geom_histogram(binwidth=1) + facet_wrap(~ AREA , scales = 'free_y')
print(fw1_free_y)
```


## *facet_wrap*: free y scale
```{r facet_wrap_4, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fw1_col <- ggplot(data=df_data, aes(x=DURATION_MINUTES)) +
  geom_histogram(binwidth=1) + 
  facet_wrap(~ AREA, ncol = 1, nrow = 2, scales = 'fixed')
print(fw1_col)
```


## *facet_wrap*: free y scale
```{r facet_wrap_5, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fw2 <- ggplot(data=df_data, aes(x=DURATION_MINUTES)) +
  geom_histogram(binwidth=1) + facet_wrap(~ AREA + Year_fac)
print(fw2)
```


## *facet_wrap*: free y scale
```{r facet_wrap_6, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fw2_2 <- ggplot(data=df_data, aes(x=DURATION_MINUTES)) +
  geom_histogram(binwidth=1) + facet_wrap( ~ Year_fac + AREA)
print(fw2_2)
```

## *facet_grid*: more flexible
```{r facet_grid_1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg1_1 <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish, color=AREA)) + 
  geom_point(shape=20) + facet_grid(. ~ Year)
print(fg1_1)
```


## *facet_grid*: change faceting display
```{r facet_grid_2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg1_2 <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish, color=AREA)) + 
  geom_point(shape=20) + facet_grid(Year ~ .)
print(fg1_2)
```



## *facet_grid*: facet with all the data
```{r facet_grid_3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg1_3 <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish, color=AREA)) + 
  geom_point(shape=20) + facet_grid(Year ~ ., margins = TRUE)
print(fg1_3)
```


## *facet_grid*: two faceting factors
```{r facet_grid_4, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg2_1 <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish)) + 
  geom_point(shape=20) + facet_grid(~ Year + AREA)
print(fg2_1)
```


## *facet_grid*: two faceting factors
```{r facet_grid_5, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg2_2 <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish)) + 
  geom_point(shape=20) + facet_grid(AREA ~ Year)
print(fg2_2)
```




## *facet_grid*: scales and space free
```{r facet_grid_6, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg2_3 <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish)) + 
  geom_point(shape=20, size=1) +
  facet_grid(AREA ~ Year, scales='free', space = 'free')
print(fg2_3)
```


## *facet_grid*: renaming labels
Replace manually names factor:
```{r facet_grid_7, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
levels(df_data$AREA) <- c('Northern area', 'Southern area')
```
Or write a function:
```{r facet_grid_8, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
## string is the levels of a factor
fn_alphabetic_label <- function(string){
  for ( i in 1:length(string)){
    string[i] <- letters[i]
  }
  return(string) 
}
```



## *facet_grid*: renaming labels
```{r facet_grid_9, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg1_3a <- ggplot(df_data, aes(x=Avg_net_depth, y=nFish, color=AREA)) + 
  geom_point(shape=20, size=1) +
  facet_grid(Year ~ ., labeller = labeller(Year = fn_alphabetic_label))
print(fg1_3a)
```




## *facet_grid*: changing facets
```{r facet_grid_10, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
fg2_3b <- fg2_2 + 
  theme(strip.text.x = element_text(size=8, angle=45),
        strip.text.y = element_text(size=12, face="bold"),
        strip.background = element_rect(colour="red", fill=NA))
print(fg2_3b)
```


## Multiple (different) plots
It can be hard to plot on the same page different plots.

For example, it is not possible to plot on the same page a boxplot, a scatter plot, and a bar plot.

```{r multi_plot, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
source('R/Fun/gg_multiplot.R')
gg_multiplot(plotlist = list(sp_c_wanderson, boxplot_TS_AREA,
                             TSplot_i95))
```


## Multiple (different) plots
```{r multi_plot2, eval=TRUE, echo=FALSE,  tidy=FALSE, opts.label='fig.large'}
gg_multiplot(plotlist = list(sp_c_wanderson, boxplot_TS_AREA,
                             TSplot_i95))
```


## Multiple (different) plots
Choose the layout to 
```{r multi_plot3, eval=FALSE,  tidy=FALSE, opts.label='fig.large'}
gg_multiplot(plotlist = list(sp_c_wanderson + ggtitle('(a)'),
                             boxplot_TS_AREA + ggtitle('(b)'),
                             TSplot_i95 + ggtitle('(c)')),
             layout = matrix(c(1,2,3,3), nrow=2, byrow=TRUE))
```


## Multiple (different) plots
```{r multi_plot4, eval=TRUE, echo=FALSE,  tidy=FALSE, opts.label='fig.large'}
gg_multiplot(plotlist = list(sp_c_wanderson + ggtitle('(a)'),
                             boxplot_TS_AREA + ggtitle('(b)'),
                             TSplot_i95 + ggtitle('(c)')),
             layout = matrix(c(1,2,3,3), nrow=2, byrow=TRUE))
```

## Theme

- ggplot2 uses a theme which controls the background, fonts, text size, ...
- change the theme for the current R session: 
```{r  eval=FALSE} 
theme_set(theme_gray())
```
- create or update your own theme :
```{r  eval=FALSE} 
p + theme(plot.background = element_rect(fill = "darkblue"))
```




## Theme: background change
```{r theme_1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
print(fg2_2 + 
        theme(plot.background = element_rect(fill = "darkblue")))
```


## Theme: black and white
```{r theme_2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
print(fg2_2 + theme_bw())
```


## Theme: classic
```{r theme_3, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
print(fg2_2 + theme_classic())
```


## Theme: minimalist
```{r theme_4, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
print(fg2_2 +  theme_minimal())
```



## Theme: minimalist tufte
```{r theme_5, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
library(ggthemes)
print(fg2_2 +  theme_tufte())
```



## Theme: the economist
```{r theme_6, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
library(ggthemes)
print(fg2_2 + theme_economist())
```


## Theme: make your own
```{r theme_7, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
theme_update(axis.text = element_text(size = rel(0.8)),
             axis.ticks = element_line(colour = "black", size=1),
             axis.line = element_line(),
             panel.background = element_rect(fill = NA, colour = NA),
             panel.border = element_rect(fill = NA, colour = NA),
             panel.grid.major = element_line(colour = "grey90", size = 0.2),
             panel.grid.minor = element_line(colour = NA, size = 0.5),
             strip.background = element_rect(fill = "grey80",colour = "grey50"),
             strip.background = element_rect(fill = "grey80",colour = "grey50"),
             panel.background = element_rect(fill = "transparent", colour = NA),
             panel.grid.minor = element_blank(),
             plot.background = element_rect(fill = "transparent", colour = NA),
             legend.key = element_rect(colour = "grey80", fill= "grey80"),
             legend.background = element_rect(colour = 'black', fill = NA),  
             legend.key = element_rect(colour = NA,fill=NA),
             legend.key.size = unit(1.1,'line'),
             legend.title = element_text(size = rel(1), face = 'bold'),
             legend.text = element_text(size = rel(1)))
```


## Legend position: top, bottom, left, right
```{r legend_p1, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend <- sp_shape_imp2 + theme(legend.position="bottom", 
                                   legend.box = "horizontal")
print(sp_legend)
```


## Legend position: bottom left *c(0,0)*, top right *c(1,1)*
```{r legend_p2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend <- sp_shape_imp2 + theme(legend.position=c(0.9,0.6))
print(sp_legend)
```


## Legend background
```{r legend_b, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend2 <- sp_legend  + 
  theme(legend.background = element_rect(fill="lightblue",
                                         size=0.5, linetype="solid", 
                                         colour ="darkblue"))
print(sp_legend2)
```



## Legend order
```{r legend_order, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend2 <- sp_legend2  + 
  guides(shape = guide_legend(order=1),
         color = guide_legend(order=2))
print(sp_legend2)
```


## Remove legends
```{r legend_rm, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend2 <- sp_legend2  + 
  theme(legend.position='none')
print(sp_legend2)
```


## Remove a specific legend
```{r legend_rm2, eval=TRUE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend_c <- sp_legend + 
  guides(color = FALSE)
print(sp_legend_c)
```


## Saving a plot
The ggsave function from ggplot2 package allows to automatically save a plot.
```{r ggsave1, eval=FALSE,  tidy=FALSE, opts.label='fig.medium'}
sp_legend_c <- sp_legend + 
  guides(color = FALSE)
print(sp_legend_c)

ggsave(plot=sp_legend_c, filename = 'sp_legend_c.pdf', path = '~/figure',
       width = 8, height = 5, units = 'cm')
```

## Useful R packages which use ggplot2

-	A website which references ggplot2 extension <https://ggplot2-exts.github.io/index.html>

-	**ggfortify** and its autoplot() function allows plotting some popular R packages using a standardized approach.
Diagnostic plots with Generalized Linear Models (GLM), Plotting Principal Component Analysis ...
- **ggmcmc** was developed to plot MCMC outputs from OpenBUGS, JAGS or Stan. It provides plots to assess the convergence and general behaviour of MCMC chains.
- **GGally** regroups R packages which extend ggplot2. The main reason to use GGally is its correlation plots.
- **latex2exp** allows to replace labels and titles with latex expression in plot.

